Verified Nanopasses for Compiling Conditionals

### 2025-06-19 paper.tex
check

@@

@@

My first exposure to these ideas were in Kent Dybvig's compiler
course~\citep{Dybvig:2010aa} where he taught the techniques used in
the Chez Scheme compiler~\citep{Dybvig:2006aa}.  One of those
techniques is called destintation-driven code
generation~\cite{Dybvig:1990aa}, which adds an extra parameter to the
compilation function to indicate how the context will use the result
value of an expression. The context may (1) ignore the result (and
only care about the expression's side effects), (2) branch based on
interpreting its result as a boolean, (3) assign the result to a
variable, or (4) output the result. These options are encoded as a
disjoint union.  Relating this to Olivier's transformation functions,
the $\mathcal{E}$ function is for contexts that need the actual value
and the $\mathcal{B}$ function is for contexts that branch on the
expression's result. Of course, a pair of functions is isomorphic to a
function on a disjoint union.
\[
(A \to C) \times (B \to C) \cong (A + B) \to C
\]

@@

@@

>>>>

\usepackage{mathpartir}
\usepackage{listings}
\usepackage{stmaryrd}

\usepackage[utf8]{inputenc}
%\usepackage[T1]{fontenc}

%\usepackage[verbose]{newunicodechar}
%% \newunicodechar{ℕ}{\ensuremath{\mathbb{N}~}}
%% \newunicodechar{𝔹}{\ensuremath{\mathbb{B}~}}
%\newunicodechar{ℤ}{\ensuremath{\mathbb{Z}~}}
%\newunicodechar{→}{\ensuremath{~\to~}}
%\newunicodechar{×}{\ensuremath{~\times~}}

\newcommand{\LangIf}{$\Lang_{\mathsf{If}}$} %R2
\newcommand{\RCO}{\mathit{mon}} % output of remove-complex-opera*
\newcommand{\LangIfANF}{\ensuremath{\Lang^{\RCO}_{\mathsf{if}}}} %R2
\newcommand{\CLang}{\mathcal{C}}
\newcommand{\LangCIf}{$\CLang_{\mathsf{If}}$} %C1

@@

\lstset{%
language=Python,
basicstyle=\ttfamily\small,
morekeywords={match,case,bool,int,let,begin,if,else,closure},
deletekeywords={},
escapechar=|,
mathescape=true,
columns=flexible,
%moredelim=[is][\color{red}]{~}{~},
showstringspaces=false,
%inputencoding=utf8
literate={→}{{\to}}1,
}

@@

My first exposure to these ideas were in Kent Dybvig's compiler
course~\citep{Dybvig:2010aa} where he taught the techniques used in
the Chez Scheme compiler~\citep{Dybvig:2006aa}.  One of those
techniques is called destintation-driven code
generation~\cite{Dybvig:1990aa}, which adds an extra parameter to the
compilation function to indicate how the context will use the result
value of an expression. The context may (1) ignore the result (and
only care about the expression's side effects), (2) branch based on
interpreting its result as a boolean, (3) assign the result to a
variable, or (4) output the result. These options are encoded as a
disjoint union.  Relating this to Olivier's transformation functions,
the $\mathcal{E}$ function is for contexts that need the actual value
and the $\mathcal{B}$ function is for contexts that branch on the
expression's result. Of course, a pair of functions is isomorphic to a
function on a disjoint union.
\[
(A → C) × (B → C) ≅ (A + B) → C
\]

@@

TODO: roadmap

@@

\section{The Languages}

In this section we describe how we formalize the syntax and semantics
of the source language and four intermediate languages in Agda:
\begin{enumerate}
\item Integers, Booleans, \texttt{let}, and \texttt{if}
\item Monadic Normal Form
\item Imperative with Structured Control Flow
\item A C-like language with \emph{goto}
\item X86 assembly but with variables
\end{enumerate}
These languages and there semantics are quite standard. However, there
are many other intermediate languages in the literature to choose
from, so why these languages, and why in this particular order?  The
answer is that they happen to represent the images of the four
nanopasses.  So then the question becomes, why do we choose to order
the nanopasses in this way? The main thing to consider is that some
passes make later passes easier because they remove a language feature
that would have been complicated to handle. For example, the Lift
Locals pass removed \texttt{let}, and I found \texttt{let} difficult
to deal with in the proof of correctness for Explicate Control
(\texttt{let} is relatively easy to handle in the code for Explicate
Control). The other thing to consider is that some compiler passes use
certain language features in their generated code, so the passes that
remove those language features better come later in the ordering.

One of the challenges of formalizing semantics in a proof assistant is
that one needs to choose a representation of variables. There are many
options, and I've tried most of them. But just like a game of
whac-a-mole, the savings acheived in one place are lost in others.  My
current favorite are good old de Bruijn indices. The following type
\lstinline{Id} is for identifiers.

\begin{lstlisting}
Id = $ℕ$
\end{lstlisting}

\subsection{The source language}

The source language is a language with integers, Booleans,
\texttt{Let} binding, and conditional expressions. We choose a couple
representative primitive operators: \texttt{Sub} for integer
subtraction and \texttt{Eq} for equality on integers and
Booleans.\footnote{Adding more primitive operators is straightforward
but uninteresting.}  The \texttt{Read} operation requests an integer
input from the user of the program.

\begin{lstlisting}
data Exp : Set where
  Num : $ℤ$ $→$ Exp
  Bool : $𝔹$ $→$ Exp
  Read : Exp
  Sub : Exp $→$ Exp $→$ Exp
  Eq : Exp $→$ Exp $→$ Exp
  Var : Id $→$ Exp
  Let : Exp $→$ Exp $→$ Exp
  If : Exp $→$ Exp $→$ Exp $→$ Exp
\end{lstlisting}

Towards specifying the semantics of this language, the inputs to the
program are an infinite stream of integers represented as a natural
number and a function from the naturals to the integers.  The natural
number is incremented each time the program reads an input.

\begin{lstlisting}
Inputs = $ℕ$ $×$ ($ℕ$ $→$ $ℤ$)
\end{lstlisting}

The following monad is used to thread the inputs through the
interpreter and to account for partiality via \lstinline{Maybe}
(e.g. looking up an undefined variable).

\begin{lstlisting}
Reader A = Inputs $→$ Maybe (A $×$ Inputs)
\end{lstlisting}

\noindent The result of an expression is an integer or Boolean, for which we
define the \lstinline{Value} type.

\begin{lstlisting}
data Value : Set where
  Int : $ℤ$ $→$ Value
  Bool : $𝔹$ $→$ Value
\end{lstlisting}

\noindent The variables (de Bruijn indices) are interpreted using an
environment that is represented as a list of values.

\begin{lstlisting}
Env A = List A
\end{lstlisting}

\noindent The semantics of expressions is given by the
\lstinline{interp-exp} function, which maps an expression and
environment to a value, in our \lstinline{Reader} monad.

\begin{lstlisting}
interp-exp : Exp $→$ Env Value $→$ Reader Value
\end{lstlisting}

\noindent The semantics of a whole program is given by the
\lstinline{interp-LIf} function.

\begin{lstlisting}
interp-LIf : Exp $→$ Inputs $→$ Maybe Value
\end{lstlisting}

\subsection{Monadic normal form}

The first intermediate language restricts the arguments of the
primitives \lstinline{Sub} and \lstinline{Eq} to be \emph{atomic}
expressions, which we define to include variables and literals. This
restriction gets us closer to x86, where instruction arguments may
only be immediates (literals) or registers.

\begin{lstlisting}
data Atm : Set where
  Num : $ℤ$ $→$ Atm 
  Bool : $𝔹$ $→$ Atm 
  Var : Id $→$ Atm

data Mon : Set where
  Atom : Atm $→$ Mon
  Read : Mon
  Sub : Atm $→$ Atm $→$ Mon
  Eq : Atm $→$ Atm $→$ Mon
  Let : Mon $→$ Mon $→$ Mon
  If : Mon $→$ Mon $→$ Mon $→$ Mon
\end{lstlisting}

The interpreter for atoms does not make use of the \lstinline{Reader}
monad because atoms do not have side effects.

\begin{lstlisting}
interp-atm : Atm $→$ Env Value $→$ Maybe Value
\end{lstlisting}

The interpreter for \lstinline{Mon} is similar to the
one for \lstinline{Exp}, except that it dispatches
to \lstinline{interp-atm} for atoms. The semantics
of a whole program is given by \lstinline{interp-LMonIf}.

\begin{lstlisting}
interp-mon : Mon $→$ Env Value $→$ Reader Value
interp-LMonIf : Mon $→$ Inputs $→$ Maybe Value
\end{lstlisting}

\subsection{Structured imperative IR}

The third language replaces the \lstinline{Let} bindings with
assignment expressions but retains the structured control-flow and
expression-oriented nature of the language. This language brings us
closer to x86 because variables are represented by registers and stack
locations, all of which are mutable. The abstract syntax for
assignment is \lstinline{(Assign $x$ $e_1$ $e_2$)} where $e_1$ is the
right-hand side and $e_2$ is an expression to be executed after the
assignment (reminiscent of the body of a \lstinline{Let}).  The
\lstinline{Program} AST node consists of the number of variables and
an expression.

\begin{lstlisting}
data Imp-Exp : Set where
  Atom : Atm $→$ Imp-Exp
  Read : Imp-Exp
  Sub : Atm $→$ Atm $→$ Imp-Exp
  Eq : Atm $→$ Atm $→$ Imp-Exp
  Assign : Id $→$ Imp-Exp $→$ Imp-Exp $→$ Imp-Exp
  If : Imp-Exp $→$ Imp-Exp $→$ Imp-Exp $→$ Imp-Exp

data Imp-Prog : Set where
  Program : $ℕ$ $→$ Imp-Exp $→$ Imp-Prog
\end{lstlisting}

Regarding the semantics, the main change is that environments are now
both an input and output. Also, in anticipation of the next language,
we switch to representing the semantics as a relation, that is, a
big-step semantics.

\begin{lstlisting}
StateImp = Inputs × Env Value
data _$⊢$_$⇓$_$⊣$_ : StateImp $→$ Imp-Exp
                     $→$ Value $→$ StateImp $→$ Set
\end{lstlisting}

\noindent Here is the rule for assignment:
\[
\inferrule{s , ρ ⊢ e_1 ⇓ v_1 ⊣ s′ , ρ′ \\
           s′ , ρ′(x := v_1) ⊣ e_2 ⇓ s″ , ρ″}
          {s , ρ ⊢ \mathtt{Assign}~ x~ e_1~ e_2 ⊣ s″ , ρ″}
\]

The semantics of whole programs is given by the following relation.

\begin{lstlisting}
interp-imp : Imp-Prog $→$ Inputs $→$ Value $→$ Set
\end{lstlisting}

\subsection{The C-like language}

The fourth language is similar to the C programming language in that
it supports \texttt{goto} and separates expressions from statements.
However, to get us closer to x86, the condition of an \lstinline{if}
statement is restricted to an equality test. This form of
\lstinline{if} can be easily translated into a sequence of three x86
instructions: \lstinline{cmpq}, \lstinline{je}, and \lstinline{jmp}.
However, compiling from the previous IR to this one is non-trivial
because the previous IR allowed arbitrarily nested expressions to
appear in the condition of an \lstinline{if}, including other
\lstinline{if}'s!

\begin{lstlisting}
data CExp : Set where
  Atom : Atm $→$ CExp
  Read : CExp
  Sub : Atm $→$ Atm $→$ CExp
  Eq : Atm $→$ Atm $→$ CExp

data CStmt : Set where
  Return : CExp $→$ CStmt
  Assign : Id $→$ CExp $→$ CStmt $→$ CStmt
  IfEq : Atm $→$ Atm $→$ Id $→$ Id $→$ CStmt
  Goto : Id $→$ CStmt

data CProg : Set where
  Program : $ℕ$ $→$ Id $→$ List CStmt $→$ CProg
\end{lstlisting}

Note that the \lstinline{CStmt} type represents not just one statement
but a block of statements that always end in either a
\lstinline{Return} statement (that halts the program) or a jump to
another block (\lstinline{Goto} or \lstinline{IfEq}).

We choose a big-step semantics for C-like language because the
semantics is no longer structurally recursive and furthermore,
programs are no longer guaranteed to terminate.\footnote{The programs
in the image of Explicate Control are guaranteed to terminate, but
Agda doesn't know that.}  So encoding the semantics as a function in
Agda would come with complications (such as using a gas parameter)
that can be side-stepped by using a big-step semantics. Compared to
the semantics for the imperative IR, this semantics refers to a list
of blocks $B$ that can be the targets of gotos.

\begin{lstlisting}
data _,_$⊢ᶜ$_$⇓$_$⊣$_ : StateImp $→$ List CStmt $→$ CStmt
                      $→$ Value $→$ StateImp $→$ Set
\end{lstlisting}

The rule for \lstinline{Goto} looks up the block for label $l$ and
executes that block.
\[
\inferrule{s , B ⊢ᶜ B[l] ⇓ v ⊣ s′}
          {s , B ⊢ᶜ \mathtt{Goto}~ l ⇓ v ⊣ s′}
\]
\noindent The rule for \lstinline{IfEq}, in the case when the
arguments are equal, is as follows:
\[
\inferrule{\texttt{interp-CExp}(\mathtt{Eq}~a_1~a_2) = \mathsf{just}~\mathsf{true} \\
           s , B ⊢ᶜ B[l_1] ⇓ v ⊣ s′}
          {s , B ⊢ᶜ \mathtt{IfEq}~a_1~a_2~l_1~l_2 ⇓ v ⊣ s′}
\]

\noindent The semantics of whole programs is given by the following
relation.

\begin{lstlisting}
interp-prog : CProg $→$ Inputs $→$ Value $→$ Set
\end{lstlisting}

\subsection{x86 with variables}

The final intermediate language is a subset of x86 assembly language,
except for the addition of variables. A complete compiler would
include a few more passes to perform register allocation and deal with
idiosyncrosies of x86, such as only allowing at most one instruction
argument to access memory.

\begin{lstlisting}
data Arg : Set where
  Num : $ℤ$ $→$ Arg
  Bool : $𝔹$ $→$ Arg
  Var : Id $→$ Arg
  Reg : $ℕ$ $→$ Arg

data Dest : Set where
  Var : Id $→$ Dest
  Reg : $ℕ$ $→$ Dest

data Inst : Set where
  MovQ : Arg $→$ Dest $→$ Inst
  SubQ : Arg $→$ Dest $→$ Inst
  CmpQ : Arg $→$ Arg $→$ Inst
  Jmp : Id $→$ Inst
  JmpEq : Id $→$ Inst
  ReadInt : Inst

Block : Set
Block = List Inst

data X86Var : Set where
  Program : $ℕ$ $→$ Id $→$ List Block $→$ X86Var
\end{lstlisting}

The mutable state consists of the input, the registers, and the
variables. However, we only use one register, \lstinline{rax}.

%input + registers + variables
\begin{lstlisting}
StateX86 = Inputs × List Value × List Value
\end{lstlisting}

The meaning of each instruction is given by the following relation.
The Boolean in this relation specifies whether the instruction
falls-through to the next instruction or not (because it jumped).

\begin{lstlisting}
data _,_$⊢$_$⇓$_,_ : StateX86 $→$ List Block $→$ Inst
                    $→$ StateX86 $→$ $𝔹$ $→$ Set
\end{lstlisting}

\noindent We take a few liberties to simplify the semantics with
respect to x86 assembly language. For example, our \lstinline{CmpQ}
instruction puts the result of the comparison into \lstinline{rax}
instead of the special \lstinline{EFLAGS} register. Likewise,
\lstinline{JmpEq} does not check the \lstinline{EFLAGS} register, but
instead jumps if \lstinline{rax} contains \lstinline{true}.

\noindent The meaning of a sequence of instructions is given the
following big-step relation.

\begin{lstlisting}
data _,_$⊩$_$⇓$_,_ : StateX86 $→$ List Block $→$ List Inst
                     $→$ StateX86 $→$ $𝔹$ $→$ Set 
\end{lstlisting}

This big-step treatment of the semantics of instruction sequences is
unusual, so here are the three rules.

\begin{gather*}
\inferrule{}{s , B ⊩ [] ⇓ s , \mathsf{true}} \\[1ex]
\inferrule{s , B ⊢ i ⇓ s′ , \mathsf{true} \\
           s′ , B ⊩ is ⇓ s″ , b}
          {s , B ⊩ i ∷ is ⇓ s′ , b}\\[1ex]
\inferrule{s , B ⊢ i ⇓ s′ , \mathsf{false}}
          {s , B ⊩ i ∷ is ⇓ s′ , b}
\end{gather*}

The meaning of a whole program is given by the following relation.
\begin{lstlisting}
interp-x86-var : X86Var → Inputs → Value → Set
\end{lstlisting}

We now proceed to discuss each compiler pass and its proof of
correctness.

\section{Pass: Remove Complex Operands}

\section{Pass: Lift Locals}

\section{Pass: Explicate Control}

\section{Pass: Select Instructions}

\section{The Compiler is Correct}
